Cache Oblivious B-trees

### 2011-12-09 paper.tex
Updating the abstract and introduction.

This paper present a survey on cache-oblivious B-trees (CO B-trees). The CO B-tree
was first introduce by Bender et al in 2000. It is an external memory data structure meaning
that it optimizes the number of memory transfers if performs rather than the number of
CPU instructions that it executions. Unlike other external memory data structure, like
the widely use B-tree, the CO B-tree does not require any information about the
number of level in the machine's memory hierarchy, the block size and capacity at each level, or the 
relative speeds for memory transfer. Hence, the CO B-tree is ``cache-oblivious''.
Despite this, the CO B-tree is able to achieve the same
optimal search bound as the B-tree and near optimal insertion and deletion bounds as well.
In this paper we first describe the original CO B-tree and explain how it is able to achieve 
its memory performance. We then present and analyze three followup CO B-tree data structures
 that improved upon shortcoming of the original: the cache-oblivious string B-tree, which introduces support 
for variable length keys, the cache-oblivious streaming B-tree, which incorporates improvements to the
insertion bound at the expense of the search bound, and lastly, the cache-oblivious concurrent B-tree, which 
introduces support for concurrency.

@@

Most algorithms assume an idealized model of computation that consists of a
single CPU, which performs instructions, and memory, which is a flat array
that can be read and written. However, in reality memory is not a flat array
but rather a hierarchy from super fast on chip caches, to slower main memory,
and even all the way down to the disk. When a program requests data, the data
is found somewhere in the memory hierarchy and typically carried up the
hierarchy to the fastest memory possible layer when the CPU is able to access
it. Unfortunately, this fast memory often has small capacity and once it is
filled and new data is requests some old entries must be evicted. Bring data
from lower to higher levels in the memory hierarchy often takes a long time.
In fact, access times between the different layers of the hierarchy can differ
by many orders of magnitude and often dominate the overall runtime performance
of an algorithm. External memory algorithms are wary of this fact and aim to
solve problem by optimizing the number of memory transfers it performs rather
than the number of instructions it executes. They often try to exploit the
fact the memory hierarchies transfer data in blocks rather than individual
atomic units. Thus, external memory algorithms are designed to not only limit
the number of memory transfers but also preserve the locality of data so that
when a memory access is performed, its cost is amortized.

@@

Since External memory algorithms are primarily concerned with efficient access
to data, it is not surprise that most problems can be reduced to finding a
suitable data structure that efficiently implement certain operations. A well
known external memory data structure is the B-tree. At a high level, B-tree
work by making their nodes the same size as a memory block in the hierarchy.
Furthermore, the fanout in the tree (the number of children) is also
proportional to the block size as well. The B-tree incurs $O(log(N))$ memory
transfers for \Search{}, \Insert{}, and \Delete{} and is widely used in
databases and filesystems.

@@

The B-tree has two theoretical shortcomings. First, the B-tree is analyzed in
a two-level memory model called the Disk Access Machine (DAM) \cite{Aggarwal}. However,
in reality computing architectures typically have memory hierarchies that are
much larger than two. The argument for the DAM model is that in practice only
one level in the memory hierarchy tends to dominate the performance and thus a
two-level memory model is sufficient for all intents and purposes. However,
this does mean that, theoretically, B-trees can begin to incur an increasing
number of memory transfers if the dominating level in the hierarchy changes.

Additionally, the B-tree expects to know the the block size in the memory
hierarchy in advanced. This means that the parameterized in code need to be
changed for every level in the memory hierarchy which can be very difficult
and error prone. Additionally, tuning this parameter may even be impossible in
a heterogeneous computing environment.

The shortcomings of the DAM model gave rise to the cache-oblivious (CO) model
of computation. In the CO model the algorithm / data structure has no
information about the memory hierarchy. It does not know anything about the
speed of any level or the block size. At a conceptual level, the CO model
allow algorithm designers to reason about the simple two-level model like DAM
but prove results for any arbitrary memory hierarchy. In short, algorithms
that perform well in a CO model should, in theory, perform well regardless of
the block size in a memory hierarchy and thus should gracefully scale to
multi-level hierarchies as well.

@@

In \cite{BenderDemainColton} Bender et al. introduce the cache-oblivious B-tree (CO B-tree). The
CO B-tree is similar to normal the B-tree except that it does not have the
block size of levels in the memory hierarchy as explicitly parameters.

The CO B-tree is able to achieve the optimal search bound of $\Theta(log(N))$
memory transfers, the optimal scan bound of $\Theta(N/B)$ memory transfers and
efficient insertion and deletion operations. The CO B-tree was a seminal data
structure that gave rise to many others in the future.

This paper provides a survey of these different data structures. We start by
analyzing the original CO B-tree and explain how it works. This includes a
detailed description of the packed memory array (PMA) and strongly
weight-balance binary search trees (SWSBT) that are integral to the CO
B-tree's performance. Following this discussion we provide an analysis of the
CO String B-tree,a CO B-tree with variable length keys, the CO Streaming
B-tree, a CO B-tree that improves the cost of insertion at the expense of
search, and the CO Concurrent B-tree (which allows multiple processes to
access the data at the same time).

>>>>

This paper present a survey on cache-oblivious B-trees (CO B-trees). The CO
B-tree was first introduce by Bender et al in 2000. It is an external memory
data structure - meaning that it optimizes the number of memory transfers it
performs rather than the number of CPU instructions that it executions. Unlike
other external memory data structures, like the widely use B-tree, the CO
B-tree does not require any information about the number of level in the
machine's memory hierarchy, the block size and capacity at each level, or the
relative speeds at each level. Hence, the CO B-tree is ``cache-oblivious''.
Despite this, the CO B-tree is able to achieve the same optimal search bound
as the B-tree and near optimal insertion and deletion bounds as well. In this
paper we first describe the original CO B-tree and explain how it is able to
achieve its memory performance. We then present and analyze three followup CO
B-tree data structures that improved upon shortcoming of the original: the
cache-oblivious string B-tree, which introduces support for variable length
keys, the cache-oblivious streaming B-tree, which incorporates improvements to
the insertion bound at the expense of the search bound, and lastly, the
cache-oblivious concurrent B-tree, which introduces support for concurrency.

@@

Most algorithms assume an idealized model of computation that consists of a
single CPU, which performs instructions, and memory, which is a flat array
that can be read and written. However, in reality memory is not a flat array
but rather a hierarchy from super fast on chip caches, to slower main memory,
to even slower disks. When a program requests data, the data is found in the
memory hierarchy and typically carried up the hierarchy to the fastest memory
layer possible, usually a cache. The CPU is then able to access it to perform
any necessary operations. Unfortunately, this cache often has a small capacity
and once it is filled any new data requests cause some existing cached data to
be evicted. Transfering data from lower to higher levels in the memory
hierarchy often takes a long time. In fact, access times between the different
layers of the hierarchy can differ by many orders of magnitude and thus the
time to transfer data often dominates the overall runtime performance of an
algorithm. External memory algorithms are wary of this fact and aim to
optimizing the number of memory transfers it performs rather than the number
of instructions it executes. They often try to exploit the fact the memory
hierarchies transfer data in blocks rather than individual atomic units. Thus,
external memory algorithms are designed to not only limit the number of memory
transfers it performs but also preserve the locality of related data so that
when a memory access is performed, its cost is amortized.

@@

Since external memory algorithms are primarily concerned with accessing data,
it is no surprise that many external memory problems can be reduced to finding
a suitable data structure that efficiently implement certain operations. A
well known external memory data structure is the B-tree. At a high level, the
B-tree works by making each one of its nodes the same size as a memory block
in a particular level of the hierarchy. Furthermore, the fanout in the tree
(the number of children of a node) is proportional to the block size as well.
The B-tree incurs $O(log(N))$ memory transfers for \Search{}, \Insert{}, and
\Delete{} and is widely used in databases and filesystems.

@@

The B-tree has two theoretical shortcomings. First, the B-tree is analyzed in
a two-level memory model called the Disk Access Machine (DAM) \cite{Aggarwal}.
However, in reality, computing architectures typically have memory hierarchies
that are much larger than two. The argument for the DAM model is that in
practice only one level in the memory hierarchy tends to dominate the
performance and thus a two-level memory model is sufficient for all intents
and purposes. However, this does mean that, theoretically, B-trees can begin
to incur an increasing number of memory transfers if the dominating level in
the hierarchy changes.

Additionally, the B-tree expects to know the block size of its memory layer in
advanced. B-Trees are implemented with parameterized code that needs to be
changed for every level in the memory hierarchy. Needless to say, this tuning
can be very difficult and error prone. Additionally, tuning this parameter may
even be impossible in certain heterogeneous computing environments.

The shortcomings of the DAM model gave rise to the cache-oblivious (CO) model
of computation. In the CO model an algorithm has no information about the
memory hierarchy. It does not know anything about the speed or the block size
of any level. Conceptually, the CO model allow algorithm designers to reason
about the simple two-level model like DAM but prove results for any arbitrary
multi-level memory hierarchy. This is because algorithms that perform well in
a CO model should, in theory, perform well regardless of changes in block
size, cache capacity, and cache speed. Consequently, CO algorithm should
gracefully scale to multi-level hierarchies as well.

@@

In \cite{BenderDemainColton} Bender et al. introduce the cache-oblivious
B-tree (CO B-tree). The CO B-tree is similar to the normal B-tree except that
it does not require explicit parameters about the memory hierarch.

The CO B-tree is able to achieve the optimal search bound of $\Theta(log(N))$
memory transfers, the optimal scan bound of $\Theta(N/B)$ memory transfers and
near optimal insertion and deletion bound of $\Theta(log_B N +
\frac{log^2{N}}{B})$. The CO B-tree was a seminal data structure that gave
rise to many others.

This paper provides a survey of the different types of CO B-trees and
discusses the various trade offs that they make. We start by analyzing the
original CO B-tree and explain how it works. This includes a detailed
description of the packed memory array (PMA) and strongly weight-balance
binary search trees (SWSBT) that are integral to the CO B-tree's memory
performance. Following this discussion we provide an analysis of the CO String
B-tree,a CO B-tree with variable length keys, the CO Streaming B-tree, a CO
B-tree that improves the cost of insertion at the expense of search, and the
CO Concurrent B-tree, which allows multiple processes to access the data at
the same time.
